\large{\bf Looking forwards and backwards through locally regulated populations }

### 2022-04-08 forwards-and-backwards.tex
lineage heuristics

\plr{
    TODO: Heuristics for how lineages should move given the population process.
}

>>>>

Although our proof of Theorem\ref{thm:lineages}
uses the explicit representation in terms of the lookdown process,
informal calculations agree with the main result.
Suppose that we are tracking a lineage that is at location $y$ at time $t$.
Looking back through time, the lineage will move at the time that individual was born,
and will jump to the location of the parent of that individual.
Now, the rate at which new individuals are born to parents at $x$ and establish at $y$
is
$$
    \gamma(x, \eta_t) q(x, dy) r(y, \eta_t) .
$$
This implies that the probability a randomly chosen individual near $y$
is a new offspring from a parent at $x$ in $[t, t+dt)$ is, informally,
\begin{equation} \label{eqn:informal_rates}
\frac{
    \gamma(x, \eta_t) q_\theta(x, dy) r(y, \eta_t) \eta_t(dx)
}{
    \eta_t(dy)
} dt .
Leaving aside questions of whether a lineage can be treated as a randomly chosen individual,
we might then proceed by defining a continuous-time jump process
whose transition rates are given by \eqref{eqn:informal_rates} given $(\eta_t)_{t=0}^T$.
Happily, we are interested in the case when $\eta$ has a density,
say $\eta_t(dx) = \varphi(x) dx$,
so interpreting these rates is not too difficult.
Now recall that $q_\theta(x, dy)$ is a Gaussian kernel
with mean $x + \meanq/\theta$ and covariance $\covq/\theta$.
Write $\widetilde q_\theta(y, dx)$ for the ``reverse-time'' distribution,
i.e., Gaussian with mean $x - \meanq/\theta$ and the same covariance.
Then, rearranging infinitesimals,
\eqref{eqn:informal_rates} implies a rate of
\begin{equation} \label{eqn:informal_rates2}
\frac{
    \varphi(x) \gamma(x, \eta_t) r(y, \eta_t)
}{
    \varphi(y)
} \widetilde q_\theta(y, dx) .
Let $(L^\theta_s)_{s=0}^T$ be the location of a lineage that moves according to
jump rates~\eqref{eqn:informal_rates2},
back through time (i.e., with the substitution that $s = T-t$).
Then
\begin{align} \label{eqn:lineage_generator_Edt}
    \begin{split}
    &\IE[f(L^\theta_{s+ds}) \;|\; L_s = y] - f(y)
    \\&\qquad 
    =
    ds \int \left(f(x) - f(y)\right)
    \frac{
        \varphi_{T-s}(x) \gamma(x, \eta_{T-s}) r(y, \eta_{T-s})
    }{
        \varphi_{T-s}(y)
    \widetilde q_\theta(y, dx) .
    \end{split}
Referring back to Definition~\ref{def:dispersal_generator},
recall that as $\theta \to \infty$,
$$
    \theta \int (f(x) - f(y)) \widetilde q_\theta(y, dx)
    \to
    - \sum_i \meanq_i \partial_{x_i} f(y)
    + \sum_{ij} \covq_{ij} \partial_{x_i x_j} f(y) ,
$$
which is just $\DG$ but with the first derivative terms multiplied by -1.
\comment{The $\widetilde \DG$ and $\covq \grad$ notation here is terrible; suggestions?}
Consequently,
    \theta \int (f(x) - f(y)) g(x) \widetilde q_\theta(y, dx)
    &\approx
    - g(y) \sum_i \meanq_i \partial_{x_i} f(y)
    + g(y) \sum_{ij} \covq_{ij} \partial_{x_i x_j} f(y)
    + \sum_{ij} \partial_{x_i} g(y) \covq_{ij} \partial_{x_j} f(y)  \\
    &=
    g(y) \widetilde \DG f(y)
    + (\covq \grad g) \cdot \grad f ,
if we define $\widetilde \DG$ to be the same as $\DG$ but with $-\meanq$ substituted for $\meanq$,
and $(\covq \grad g(y))_i = \sum_j \covq_{ij} \partial_{x_j} g(y)$.
Applying this to~\eqref{eqn:lineage_generator_Edt} with $g = \varphi_{T-s} \gamma$,
this suggests that the generator of the limiting process is
    \Lgen_s f
    &=
    \frac{r}{\varphi_{T-s}}
    \left\{
        \varphi_{T-s} \gamma \widetilde \DG f
        + \left(\covq \grad (\varphi_{T-s} \gamma) \right) \cdot \grad f 
    \right\} \\
    &=
    r \gamma
    \left\{
        \widetilde \DG f
        + \left(\covq \grad \log(\varphi_{T-s} \gamma) \right) \cdot \grad f 
    \right\} .
In other words, the lineage moves at speed equal to the local production of successful offspring,
$r \gamma$,
by a time-reversed version of the dispersal distribution,
but with a drift in the direction of the gradient of the logarithm
of the production of juveniles, $\grad \log(\phi_{T-s} \gamma)$
(modified by the covariance of dispersal).
This \comment{presumably?} agrees with \comment{the theorem}.
